### 2025-10-27 Text/Thesis/Frontmatter/abstract.tex
final version thesis

This work  presents a statistical approach to detect heterogeneity 
in synthetic aperture radar (SAR) intensity data using entropy-based methods.
In homogeneous regions, where fully developed speckle is present, SAR returns 
are well modeled by the Gamma distribution. 
In contrast, heterogeneous areas exhibit more complex scattering behaviors, 
typically modeled by the $\mathcal{G}_I^0$ distribution. Distinguishing between 
these regimes is fundamental for accurate interpretation in remote sensing applications, 
yet classical parametric tests often fail to address this task effectively due to analytical 
and numerical limitations.
To overcome these challenges, we propose three test statistics for detecting heterogeneity 
in SAR imagery based on Shannon, Rényi, and Tsallis entropy measures. These tests rely on 
nonparametric entropy estimators constructed from sample spacings, avoiding explicit assumptions 
about the underlying distribution. To enhance their accuracy, particularly in small samples, 
we incorporate a bootstrap bias-correction procedure that improves the estimators’ stability, 
reduces bias, and lowers mean squared error.
The proposed tests are evaluated using Monte Carlo simulations, assessing size and power 
under varying sample sizes and numbers of looks.
Results demonstrate that the Rényi and Tsallis-based tests outperform the Shannon-based version by 
detecting more subtle texture variations while maintaining greater reliability in identifying 
truly homogeneous regions.
Finally, the methodology is applied to both simulated and real SAR data. We generate $p$-value maps using a 
sliding window analysis, allowing for visual and quantitative assessment of spatial heterogeneity. 
The Rényi-based test consistently identifies fine-scale roughness patterns, 
while the Tsallis-based test shows strength in reliably detecting homogeneous areas. 
Together, these entropy-based tools provide a robust, interpretable, and unsupervised 
framework for heterogeneity detection in SAR data.

\vspace{1em}
\par
\noindent \textbf{Keywords:} SAR; Heterogeneity; Entropy; Nonparametric Estimation; Bootstrap; Statistical Test

>>>>

This work presents a statistical approach to detect heterogeneity 
in synthetic aperture radar (SAR) intensity data using entropy-based methods.
In SAR data analysis, an accurate interpretation of the terrain fundamentally 
depends on the distinction between two main regimes: homogeneous regions, in which 
speckle is fully developed and SAR returns are represented by the Gamma distribution, 
and heterogeneous areas, which require more flexible distributions to describe complex scattering, 
generally represented by the $\mathcal{G}_I^0$ distribution.
Although this discrimination is essential for remote sensing applications, 
classical parametric tests are generally not suitable for this task due to analytical and numerical limitations.
To overcome these challenges, we propose three test statistics to detect heterogeneity in SAR images based on 
Shannon, Rényi, and Tsallis entropies. 
The associated tests employ nonparametric entropy estimators constructed from sample spacings, avoiding explicit 
assumptions about the underlying distribution. To increase the accuracy of the tests, especially for small samples, 
we incorporate a bootstrap-based bias correction procedure that improves the stability of the estimators, reduces bias, 
and decreases the mean squared error.
The proposed tests are evaluated through Monte Carlo simulations, using test size and power under different speckle 
and texture conditions as performance criteria.
The results show that the tests based on Rényi and Tsallis entropies outperform the version based on Shannon entropy, 
detecting subtler texture variations and maintaining higher reliability in identifying truly homogeneous regions.
Finally, the methodology is applied to both simulated and real SAR data.
The analysis is performed using sliding windows, generating maps of $p$-values that allow for visual 
and quantitative assessment of spatial heterogeneity. 
The Rényi-based test consistently identifies fine-scale roughness patterns, while the Tsallis-based test performs better in detecting homogeneous regions. 
Together, these entropy-based tools provide a robust, interpretable, and unsupervised framework for detecting heterogeneity in SAR data.

\vspace{1em}
\par
\noindent \textbf{Keywords:}  entropy;  heterogeneity; gamma distribution; hypothesis testing; bootstrap.

### 2025-10-27 Text/Thesis/Frontmatter/acknowledgements.tex
final version thesis

Firstly, I would like to thank my supervisors, Prof. Alejandro Frery and Prof. Abraão do Nascimento, for their guidance, dedication, and support throughout these years. 
Working with you has been an incredibly valuable learning experience. 
I truly appreciate the attention and dedication you give to your students.

I am also grateful to the Universidade Federal de Pernambuco and the 
Graduate Program in Statistics for their support during my academic journey. 
Special thanks to Prof. Pablo Rodriguez for all his help with administrative matters and for his patience.

I gratefully acknowledge the financial support provided by the Coordenação de Aperfeiçoamento de Pessoal de Nível Superior – Brasil (CAPES) – Finance Code 001, 
the Programa de Doutorado-Sanduíche no Exterior (PDSE), and the Fundação de Amparo à Ciência e Tecnologia do Estado de Pernambuco (FACEPE).

I also thank the examination committee for their availability.

To my family, thank you for always supporting and encouraging me along the way.

Finally, thanks to all the friends I met during this time. Your support and friendship 
made this experience better.

>>>>

Firstly, I would like to thank my supervisors, Prof. Alejandro Frery and Prof. Abraão do Nascimento, for their guidance and support throughout these years. Working with you has been an incredibly valuable learning experience. I truly appreciate the attention and dedication you give to your students.

I am also thankful to the Universidade Federal de Pernambuco and the Programa de
Pós-Graduação em Estatística for their support during my academic journey. Special thanks to Prof. Pablo Rodriguez for all his help with administrative matters and for his patience.

I acknowledge the financial support from the Coordenação de Aperfeiçoamento de Pessoal de Nível Superior – Brasil (CAPES) – Finance Code 001, the Programa de Doutorado-Sanduíche no Exterior (PDSE), and the Fundação de Amparo à Ciência e Tecnologia do Estado de Pernambuco (FACEPE), which made this work possible.

I would like to express my gratitude to the members of my examination committee, Prof. Saeid Homayouni, Prof. Paolo Gamba, Prof. Anderson de Borba, and Prof. Fátima Sombra, for agreeing to be part of the defense of my thesis and for their valuable comments and suggestions.

This journey would not have been the same without all the amazing people I met in Brazil and New Zealand. To everyone, thank you for your friendship and support, which made this experience unforgettable.

Finally, I am especially grateful to my family, who have always supported and encouraged me along the way.

### 2025-10-27 Text/Thesis/Frontmatter/resumo.tex
final version thesis

Este trabalho apresenta uma abordagem estatística para a detecção de heterogeneidade 
em dados de intensidade de radar de abertura sintética (SAR) utilizando medidas 
baseadas em entropia. Em regiões homogêneas, onde há speckle totalmente desenvolvido, 
os retornos SAR são bem modelados pela distribuição Gama. Em contraste, áreas heterogêneas 
apresentam comportamentos de espalhamento mais complexos, normalmente descritos pela 
distribuição $\mathcal{G}_I^0$. Distinguir entre esses dois regimes é fundamental 
para uma interpretação precisa em aplicações de sensoriamento remoto. No entanto, 
testes paramétricos clássicos geralmente não são adequados para essa tarefa devido a 
limitações analíticas e numéricas.
Para superar esses desafios, propomos três estatísticas de teste para detectar heterogeneidade 
em imagens SAR com base nas entropias de Shannon, Rényi e Tsallis. Esses testes utilizam 
estimadores não paramétricos de entropia construídos a partir de espaçamentos amostrais, 
evitando suposições explícitas sobre a distribuição subjacente. Para aumentar a precisão dos testes, 
especialmente em amostras pequenas, incorporamos um procedimento de correção de viés via bootstrap, 
que melhora a estabilidade dos estimadores, reduz o viés e o erro quadrático médio.
As estatísticas propostas são avaliadas por meio de simulações de Monte Carlo, 
onde analisamos seu tamanho e poder sob diferentes condições de speckle e textura. 
Os resultados mostram que os testes baseados em Rényi e Tsallis superam a versão baseada 
em Shannon, detectando variações de textura mais sutis e mantendo maior confiabilidade na 
identificação de regiões verdadeiramente homogêneas.
Por fim, a metodologia é aplicada a dados SAR simulados e reais. 
A análise é realizada com janelas deslizantes, gerando mapas de valores de $p$ que permitem a a
valiação visual e quantitativa da heterogeneidade espacial. O teste baseado na entropia de 
Rényi mostra desempenho superior na identificação de padrões de rugosidade em pequena escala, 
enquanto o teste baseado em Tsallis é eficaz na detecção de regiões homogêneas. Em conjunto, 
essas ferramentas baseadas em entropia oferecem uma estrutura robusta, interpretável e não 
supervisionada para a detecção de heterogeneidade em dados SAR.

\vspace{1em}
\par
\noindent \textbf{Palavras-chave:} SAR; Heterogeneidade; Entropia; Estimativa não paramétrica; Bootstrap; Teste estatístico

>>>>

Este trabalho apresenta uma abordagem estatística para detectar heterogeneidade em dados de
intensidade de radar de abertura sintética (SAR), utilizando métodos baseados em entropia.
Na análise de dados SAR, uma interpretação precisa do terreno depende fundamentalmente da
distinção entre dois regimes principais: regiões homogêneas, em que o speckle está
totalmente desenvolvido, os retornos SAR são representados pela distribuição Gamma, e áreas heterogêneas requerem distribuições mais flexíveis para descrever
o espalhamento complexo, geralmente representado pela distribuição
$\mathcal{G}_I^0$.
Embora essa discriminação seja essencial para aplicações de sensoriamento remoto, 
testes paramétricos clássicos geralmente não são adequados para essa tarefa devido a 
limitações analíticas e numéricas.
Para superar esses desafios, propomos três estatísticas de teste para detectar heterogeneidade
em imagens SAR, com base nas entropias de Shannon, Rényi e Tsallis. Os testes associados utilizam
estimadores não paramétricos de entropia construídos a partir de espaçamentos amostrais,
evitando suposições explícitas sobre a distribuição subjacente. Para aumentar a precisão dos testes,
especialmente em amostras pequenas, incorporamos um procedimento de correção de viés via bootstrap,
que melhora a estabilidade dos estimadores, reduz o viés e o erro quadrático médio.
Os testes propostos são avaliados por meio de simulações de Monte Carlo,
tendo como critério de avaliação seu tamanho e poder sob diferentes condições de speckle e textura.
Os resultados mostram que os testes baseados nas entropias de Rényi e Tsallis superam a versão baseada
na entropia de Shannon, detectando variações de textura mais sutis e mantendo maior confiabilidade na
identificação de regiões verdadeiramente homogêneas.
Por fim, a metodologia é aplicada tanto a dados simulados como a dados SAR reais.
A análise é realizada com janelas deslizantes, gerando mapas de valores-$p$ que permitem a
avaliação visual e quantitativa da heterogeneidade espacial. O teste baseado na entropia de
Rényi mostra desempenho superior na identificação de padrões de rugosidade em pequena escala,
enquanto o teste baseado em Tsallis é melhor na detecção de regiões homogêneas. Em conjunto,
essas ferramentas baseadas em entropia oferecem uma estrutura robusta, interpretável e não
supervisionada para a detecção de heterogeneidade em dados SAR.

\vspace{1em}
\par
\noindent \textbf{Palavras-chave:} entropia;  heterogeneidade; distribuição gama; teste de hipótese; bootstrap.

### 2025-10-27 Text/Thesis/Frontmatter/symbols.tex
final version thesis

\begin{symbols}{lll} % Include a list of Symbols (a three column table)

$\mathbb{R}_+$ & Positive real numbers \\
 $\Gamma(\cdot)$  & Gamma function \\
$\Gamma_{\mathrm{SAR}}(L, \mu)$ & Gamma SAR distribution with parameters $L$ and $\mu$ \\
$\mathcal{G}^0_I$ & $\mathcal{G}^0_I$ distribution for intensity \\
$\psi^{(0)}(\cdot)$ & Digamma function \\
$H(Z)$ & Shannon entropy of random variable $Z$ \\
$R_{\lambda}(Z)$ & Rényi entropy of order $\lambda$ \\
$T_{\lambda}(Z)$ & Tsallis entropy of order $\lambda$ \\
$\widehat{H}_{\text{AO}}(\bm{Z})$ & Nonparametric Shannon entropy estimator (Al-Omari) \\
$\widehat{R}_{\lambda}(\bm{Z})$ & Nonparametric Rényi entropy estimator via sample spacings \\
$\widehat{T}_{\lambda}(\bm{Z})$ & Nonparametric Tsallis entropy estimator via sample spacings \\
$\widetilde{H}_{\text{AO}}(\bm{Z})$ & Bootstrap-corrected Shannon entropy estimator \\
$\widetilde{R}_{\lambda}(\bm{Z})$ & Bootstrap-corrected Rényi entropy estimator \\
$\widetilde{T}_{\lambda}(\bm{Z})$ & Bootstrap-corrected Tsallis entropy estimator \\
$S_{\widetilde{H}_{\text{AO}}}(\bm{Z}; L)$ & Test statistic based on Shannon entropy \\
$S_{\widetilde{R}_{\lambda}}(\bm{Z}; L)$ & Test statistic based on Rényi entropy \\
$S_{\widetilde{T}_{\lambda}}(\bm{Z}; L)$ & Test statistic based on Tsallis entropy \\
$\widehat{\mu}$ & Sample mean of $\bm{Z}$ \\
$Z_{(i)}$ & $i$-th order statistic in sample $\bm{Z}$ \\
$m$ & Spacing parameter (window size) used in entropy estimation \\
$c_i$ & Boundary correction factor in spacing estimators \\
\addlinespace % Gap to separate the Roman symbols from the Greek

\end{symbols}

>>>>

% --- SOLO afecta a esta tabla ---
{%
\setlength{\extrarowheight}{0.5pt}   % extra vertical por fila
\renewcommand{\arraystretch}{1.18} % factor global para altura de filas (1.0 = normal)

\begin{symbols}{@{}lll@{}} % @{} opcional: elimina márgenes laterales de la tabla
$\mathbb{R}_+$ & Positive real numbers \\
$\Gamma(\cdot)$  & Gamma function \\
$\Gamma_{\mathrm{SAR}}(L, \mu)$ & Gamma SAR distribution with parameters $L$ and $\mu$ \\
$\mathcal{G}^0_I$ & $\mathcal{G}^0_I$ distribution for intensity \\
$\psi^{(0)}(\cdot)$ & Digamma function \\
$H(Z)$ & Shannon entropy of random variable $Z$ \\
$R_{\lambda}(Z)$ & Rényi entropy of order $\lambda$ \\
$T_{\lambda}(Z)$ & Tsallis entropy of order $\lambda$ \\
$\widehat{H}_{\text{AO}}(\bm{Z})$ & Nonparametric Shannon entropy estimator (Al-Omari) \\
$\widehat{R}_{\lambda}(\bm{Z})$ & Nonparametric Rényi entropy estimator via sample spacings \\
$\widehat{T}_{\lambda}(\bm{Z})$ & Nonparametric Tsallis entropy estimator via sample spacings \\
$\widetilde{H}_{\text{AO}}(\bm{Z})$ & Bootstrap-corrected Shannon entropy estimator \\
$\widetilde{R}_{\lambda}(\bm{Z})$ & Bootstrap-corrected Rényi entropy estimator \\
$\widetilde{T}_{\lambda}(\bm{Z})$ & Bootstrap-corrected Tsallis entropy estimator \\
$S_{\widetilde{H}_{\text{AO}}}(\bm{Z}; L)$ & Test statistic based on Shannon entropy \\
$S_{\widetilde{R}_{\lambda}}(\bm{Z}; L)$ & Test statistic based on Rényi entropy \\
$S_{\widetilde{T}_{\lambda}}(\bm{Z}; L)$ & Test statistic based on Tsallis entropy \\
$\widehat{\mu}$ & Sample mean of $\bm{Z}$ \\
$Z_{(i)}$ & $i$-th order statistic in sample $\bm{Z}$ \\
$m$ & Spacing parameter (window size) used in entropy estimation \\
$c_i$ & Boundary correction factor in spacing estimators \\
\addlinespace[4pt] % (opcional) espacio extra para separar bloques
\end{symbols}
}% --- fin del ámbito local ---

% \begin{symbols}{lll} % Include a list of Symbols (a three column table)
% 
% $\mathbb{R}_+$ & Positive real numbers \\ 
%  $\Gamma(\cdot)$  & Gamma function \\
% $\Gamma_{\mathrm{SAR}}(L, \mu)$ & Gamma SAR distribution with parameters $L$ and $\mu$ \\
% $\mathcal{G}^0_I$ & $\mathcal{G}^0_I$ distribution for intensity \\
% $\psi^{(0)}(\cdot)$ & Digamma function \\
% $H(Z)$ & Shannon entropy of random variable $Z$ \\
% $R_{\lambda}(Z)$ & Rényi entropy of order $\lambda$ \\
% $T_{\lambda}(Z)$ & Tsallis entropy of order $\lambda$ \\
% $\widehat{H}_{\text{AO}}(\bm{Z})$ & Nonparametric Shannon entropy estimator (Al-Omari) \\
% $\widehat{R}_{\lambda}(\bm{Z})$ & Nonparametric Rényi entropy estimator via sample spacings \\
% $\widehat{T}_{\lambda}(\bm{Z})$ & Nonparametric Tsallis entropy estimator via sample spacings \\
% $\widetilde{H}_{\text{AO}}(\bm{Z})$ & Bootstrap-corrected Shannon entropy estimator \\
% $\widetilde{R}_{\lambda}(\bm{Z})$ & Bootstrap-corrected Rényi entropy estimator \\
% $\widetilde{T}_{\lambda}(\bm{Z})$ & Bootstrap-corrected Tsallis entropy estimator \\
% $S_{\widetilde{H}_{\text{AO}}}(\bm{Z}; L)$ & Test statistic based on Shannon entropy \\
% $S_{\widetilde{R}_{\lambda}}(\bm{Z}; L)$ & Test statistic based on Rényi entropy \\
% $S_{\widetilde{T}_{\lambda}}(\bm{Z}; L)$ & Test statistic based on Tsallis entropy \\
% $\widehat{\mu}$ & Sample mean of $\bm{Z}$ \\
% $Z_{(i)}$ & $i$-th order statistic in sample $\bm{Z}$ \\
% $m$ & Spacing parameter (window size) used in entropy estimation \\
% $c_i$ & Boundary correction factor in spacing estimators \\
% \addlinespace % Gap to separate the Roman symbols from the Greek
% 
% 
% \end{symbols}
